### 2020-03-14 thesis/content/chapter-graph-lta.tex
continued svm lta proof + minor fixes

		Note that there are no explicit constituent weights in the existing unstructured \ac{lta} approaches (i.e.\ implicitly all $w_{G, i} = 1$) because the explicitly given constituents are assumed to be equally indicative of $y_G$.
		For structured data however, where the decomposition $\psi(G)$ is not given as part of the input, this assumption does not necessarily hold.
		By weighting the constituents, an \ac{lta} model can reduce the relevance or even ignore constituents that turn out to be irrelevant for the compositions target score $y_G$.
	\item \textbf{Aggregation:}
		Lastly a \textit{weighted aggregation function} $\mathcal{A}: {(\mathcal{Y} \times \mathbb{R}_{\geq 0})}^{*} \to \mathcal{Y}$ must be applied.
		It combines the multiset of local constituent predictions and non-negative weights into a single global composition prediction.
		\begin{defn}\label{defn:ltag:weighted-agg}
			We call $\mathcal{A}$ a \textit{weighted aggregation function} iff.\ it satisfies
			\begin{align*}
				\text{idempotency: } & \exists c \geq 0: \forall y \in \mathcal{Y}, w \in \mathbb{R}_{\geq 0}^n\text{ s.t. }{\max w} \geq c: \mathcal{A}({\ldblbrace (y, w_i) \rdblbrace}_{i=1}^n) = y \\
				\land\text{ zero invar.: } & \forall y_0 \in \mathcal{Y}, S = {\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n: \mathcal{A}(S \cup \{ (y_0, 0) \}) = \mathcal{A}(S) \text{.}
			\end{align*}
		\end{defn}
		{\setlength{\parskip}{0pt}The idempotency constraint requires aggregation functions to agree with uniform input scores $y$ if at least one input is weighted above some threshold $c$.
		The zero invariance constraint requires aggregation functions to ignore inputs with zero weight.}

		One example for such a weighted aggregator is the \textit{weighted mean} function $\wmean({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq \sum_{i=1}^n w_i y_i$ which requires $\sum w_i = 1$ and $w_i \in [0, 1]$.
		Alternatively an unweighted aggregation function like $\min$, $\max$ or \ac{owa}\footnote{
			Even though \ac{owa} uses weights, it does not get those weights as part of the input and is therefore considered to be unweighted here.
		} also trivially satisfies \cref{defn:ltag:weighted-agg} if all inputs $y_i$ with $w_i = 0$ are filtered out and the weights for all remaining inputs are ignored.
\end{enumerate}
Based on the notion of decomposition, disaggregation and aggregation we can now define the concept of \textit{\ac{lta} formulations}.
\begin{defn}
	A model $h: \mathcal{G} \to \mathcal{Y}$ is in an \textit{\ac{lta} formulation} iff.\ it is expressed as
	\begin{align*}
		h(G) \coloneqq \mathcal{A}(\ldblbrace f(c_{G,i})\, |\, c_{G,i} \in \psi(G) \rdblbrace) \quad\text{with $\psi$, $f$ and $\mathcal{A}$ as defined above.}
	\end{align*}
\end{defn}
Note that every model $h: \mathcal{G} \to \mathcal{Y}$ has a trivial recursive \ac{lta} formulation by choosing $\psi(G) = \{ G \}$, $f(G) = (h(G), 1)$ and an arbitrary weighted aggregation function $\mathcal{A}$.
Those trivial \ac{lta} formulations do not split compositions into locally evaluated constituents and therefore intuitively do not fulfill the postulated localized explainability characteristic of \ac{lta}.
However since there is no commonly accepted formal criterion to decide whether a model's decisions are explainable~\cite{Lipton2018}, we do not attempt to strictly distinguish between \ac{lta} and non-\acs{lta} methods.
Instead the notion of \ac{lta} formulations should be seen as way to identify how ``\acs{lta}-like'' a model is:
\begin{itemize}
	\item \textbf{Negative extreme:}
		If a model $h$ only has trivial \ac{lta} formulations with the decomposition function $\psi(G) = \{ G \}$, it is not considered to be an \ac{lta} model.
	\item \textbf{Positive extreme:}
		If a model has an \ac{lta} formulation with a decomposition function that returns interpretable constituents, it is considered to be an \ac{lta} model.
		By definition this is true for the constituents $\psi(G) \coloneqq {\{ G[v_{c_i}] \}}_{v_{c_i} \in \mathcal{V}_G}$ of \ac{lta} methods for unstructured data.
	\item \textbf{In-between cases:}
		An \ac{lta} method for structured data must produce models that lie somewhere in-between the two extremes.
\end{itemize}

@@

In \cref{sec:related:gcr:embed,sec:related:gcr:kernel} three different ways to map a given graph $G$ to a vector $\varphi(G) \in \mathbb{R}^d$ were described:
\begin{enumerate*}
	\item Fingerprint embeddings,
	\item skip-gram inspired embeddings and
	\item kernel embeddings
\end{enumerate*}.
One common way to solve the \ac{gcr} problem via those embedding vectors is to train an \ac{svm} on them.
We will now see that \acp{svm} can be interpreted as \ac{lta} models if they are trained on so-called \textit{substructure component embeddings}.
\begin{defn}\label{defn:ltag:substruct-embedding}
	A graph embedding $\varphi: \mathcal{G} \to \mathbb{N}^{d}$ is called a \textit{substructure component embedding} iff.\ %
	$\forall i \in [d]: \exists S_i \in \mathcal{G}_{\mathit{ord}}, \gamma_i \in \mathbb{R}: \forall G \in \mathcal{G}: {{\varphi(G)}[i] = \mathit{contains}_{S_i}(G)} \lor {{\varphi(G)}[i] = \gamma_i \mathit{count}_{S_i}(G)}$,
	with $\mathit{contains}$ and $\mathit{count}$ as in \cfullref[ on ]{defn:related:wl-count-detect}.
\end{defn}
Intuitively \cref{defn:ltag:substruct-embedding} states that every embedding dimension must represent the count or presence of some ordered subgraph $S_i$ in the embedded graph $G$.
Based on this requirement we now proof the main theorem which shows the relation between \acp{svm} and \ac{lta}.
\begin{thm}
	A binary \ac{svm} classifier that takes substructure component embeddings $\varphi(G)$ as its input has an \ac{lta} formulation with a nontrivial decomposition function $\psi$.
\end{thm}
\begin{proof}
	Let $h: \mathcal{G} \to {\{ -1, +1 \}}$ be a binary graph classifier expressed as $h = h_{\text{\acs*{svm}}} \circ \varphi$ where $\varphi: \mathcal{G} \to \mathbb{N}_0^{d}$ is a substructure component embedding and $h_{\text{\acs*{svm}}}: \mathbb{R}^{d} \to {\{ -1, +1 \}}$ an \ac{svm} classifier.
	If $h$ is trained on a dataset $\mathcal{D}_{\text{train}} = {\{ (G_1, y_1), \dots, (G_N, y_N) \}}$, via the kernel trick it can be expressed as
	{\allowdisplaybreaks\begin{align*}
		h(G) &= \sgn\left( \sum_{j = 1}^{N} \alpha_j y_j {\langle \varphi(G), \varphi(G_j) \rangle} + b \right)
		\quad\text{for some $\alpha \in \mathbb{R}^N$ and $b \in \mathbb{R}$} \\
		&= \sgn\left( {\sum_{i=1}^{d} {\varphi(G)}[i] \underbrace{\left( \sum_{j = 1}^{N} \alpha_j y_j {\varphi(G_j)}[i] \right)}_{{} \eqqcolon z_i}} + b \right)
		\text{.}
	\end{align*}}

	By \cref{defn:ltag:substruct-embedding} each $\varphi(G)[i] > 0$ counts or indicates the presence of some ordered subgraph $S_i$ in $G$;
	more precisely:
	\begin{itemize}
		\item If $\varphi(G)[i] = \gamma_i \mathit{count}_{S_i}(G) > 0$, let $C_i \coloneqq {\{ G[s]\, |\, s \in \mathcal{P}(\mathcal{V}_G) \land G[s] \simeq S_i \}}$ be the $\mathit{count}_{S_i}(G)$ subgraphs counted by the $i$-th dimension.
		\item If $\varphi(G)[i] = \mathit{contains}_{S_i}(G) = 1$, let $C_i \coloneqq {\{  \}}$.
	\end{itemize}
	A substructure component embedding vector can therefore be interpreted as an encoding of the graph decomposition \[
		\psi(G) \coloneqq \bigcup_{i=1}^{d}
		\text{.}
	\]
	\begin{align*}
		h(G) &= \sgn\left( {\sum_{i=1}^{d} {\varphi(G)}[i] \underbrace{|z_i|}_{{} \eqqcolon \nu_i} \underbrace{\sgn(z_i)}_{{} \eqqcolon y_i}} + b \right)
	\end{align*}
\end{proof}

>>>>

		Note that there are no explicit constituent weights in the existing unstructured \ac{lta} approaches (i.e.\ implicitly all $w_{G, i} = 1$) because the explicitly given constituents are assumed to be equally indicative of $y_G$.
		For structured data however, where the decomposition $\psi(G)$ is not given as part of the input, this assumption does not necessarily hold.
		By weighting the constituents, an \ac{lta} model can reduce the relevance or even ignore constituents that turn out to be irrelevant for the compositions target score $y_G$.
	\item \textbf{Aggregation:}
		Lastly a \textit{weighted aggregation function} $\mathcal{A}: {(\mathcal{Y} \times \mathbb{R}_{\geq 0})}^{*} \to \mathcal{Y}$ must be applied.
		It combines the multiset of local constituent predictions and non-negative weights into a single global composition prediction.
		\begin{defn}\label{defn:ltag:weighted-agg}
			We call $\mathcal{A}$ a \textit{weighted aggregation function} iff.\ it satisfies
			\begin{align*}
				\text{idempotency: } & \exists c > 0: \forall y \in \mathcal{Y}, w \in \mathbb{R}_{\geq 0}^n\text{ s.t. }{\max w} \geq c: \mathcal{A}({\ldblbrace (y, w_i) \rdblbrace}_{i=1}^n) = y \\
				\land\text{ zero invar.: } & \forall y_0 \in \mathcal{Y}, S = {\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n: \mathcal{A}(S \cup \{ (y_0, 0) \}) = \mathcal{A}(S) \text{.}
			\end{align*}
		\end{defn}
		{\setlength{\parskip}{0pt}The idempotency constraint requires aggregation functions to agree with uniform input scores $y$ if at least one input is weighted above some threshold $c$.
		The zero invariance constraint requires aggregation functions to ignore inputs with zero weight.
		Exemplary weighted aggregation function are:}
		\begin{itemize}
			\item The \textit{weighted mean} function $\wmean({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq \sum_{i=1}^n w_i y_i$ which requires $\sum w_i = 1$, $w_i \in [0, 1]$ and the existence of some scalar multiplication operator $\cdot: \mathbb{R} \times \mathcal{Y} \to \mathcal{Y}$, typically from some subfield $\mathcal{Y} \subseteq \mathbb{R}$ or possibly also some vector subspace $\mathcal{Y} \subseteq \mathbb{R}^d$.
			\item Another basic weighted aggregator is the \textit{weighted majority vote} function $\wmaj({\ldblbrace (y_i, w_i) \rdblbrace}_{i=1}^n) \coloneqq {\arg\max}_{y \in \mathcal{Y}} \sum_{y_i = y} w_i$ which returns the input with the highest total weight.
				Unlike $\wmean$ it can also be applied to score domains $\mathcal{Y}$ without a multiplication operator, e.g.\ sets of discrete classes.
			\item Alternatively an unweighted aggregation function like $\min$, $\max$, $\mean$ or \ac{owa}\footnote{
				Even though \ac{owa} uses weights, it does not get those weights as part of the input and is therefore considered to be unweighted in this context.
			} also trivially satisfies \cref{defn:ltag:weighted-agg} if all inputs $y_i$ with $w_i = 0$ are filtered out and the weights for all remaining inputs are ignored.
		\end{itemize}
\end{enumerate}
Based on the notion of decomposition, disaggregation and aggregation we can now define the concept of \textit{\ac{lta} formulations}.
\begin{defn}
	A model $h: \mathcal{G} \to \mathcal{Y}$ is in an \textit{\ac{lta} formulation} iff.\ it is expressed as
	\begin{align*}
		h(G) \coloneqq \mathcal{A}(\ldblbrace f(c_{G,i})\, |\, c_{G,i} \in \psi(G) \rdblbrace) \quad\text{with $\psi$, $f$ and $\mathcal{A}$ as defined above.}
	\end{align*}
\end{defn}
Note that every model $h: \mathcal{G} \to \mathcal{Y}$ has a trivial recursive \ac{lta} formulation by choosing $\psi(G) = \{ G \}$, $f(G) = (h(G), 1)$ and an arbitrary weighted aggregation function $\mathcal{A}$.
Those trivial \ac{lta} formulations do not split compositions into locally evaluated constituents and therefore intuitively do not fulfill the postulated localized explainability characteristic of \ac{lta}.
However since there is no commonly accepted formal criterion to decide whether a model's decisions are explainable~\cite{Lipton2018}, we do not attempt to strictly distinguish between \ac{lta} and non-\acs{lta} methods.
Instead the notion of \ac{lta} formulations should be seen as way to identify how ``\acs{lta}-like'' a model is:
\begin{itemize}
	\item \textbf{Negative extreme:}
		If a model $h$ only has trivial \ac{lta} formulations with the decomposition function $\psi(G) = \{ G \}$, it is not considered to be an \ac{lta} model.
	\item \textbf{Positive extreme:}
		If a model has an \ac{lta} formulation with a decomposition function that returns interpretable constituents, it is considered to be an \ac{lta} model.
		By definition this is true for the constituents $\psi(G) \coloneqq {\{ G[v_{c_i}] \}}_{v_{c_i} \in \mathcal{V}_G}$ of \ac{lta} methods for unstructured data.
	\item \textbf{In-between cases:}
		An \ac{lta} method for structured data must produce models that lie somewhere in-between the two extremes.
\end{itemize}

@@

In \cref{sec:related:gcr:embed,sec:related:gcr:kernel} three different ways to map a given graph $G$ to a vector $\varphi(G) \in \mathbb{R}^d$ were described:
\begin{enumerate*}
	\item Fingerprint embeddings,
	\item skip-gram inspired embeddings and
	\item kernel embeddings
\end{enumerate*}.
One common way to solve the \ac{gcr} problem via those embedding vectors is to train an \ac{svm} on them.
We will now see that \acp{svm} can be interpreted as \ac{lta} models if they are trained on so-called \textit{substructure component embeddings}.
\begin{defn}\label{defn:ltag:substruct-embedding}
	A graph embedding $\varphi: \mathcal{G} \to \mathbb{N}^{d}$ is called a \textit{substructure component embedding} iff.\ %
	$\forall i \in [d]: \exists S_i \in \mathcal{G}_{\mathit{ord}}, \gamma_i > 0: \forall G \in \mathcal{G}: {{\varphi(G)}[i] = \gamma_i \mathit{contains}_{S_i}(G)} \lor {{\varphi(G)}[i] = \gamma_i \mathit{count}_{S_i}(G)}$,
	with $\mathit{contains}$ and $\mathit{count}$ as in \cfullref[ on ]{defn:related:wl-count-detect}.
\end{defn}
Intuitively \cref{defn:ltag:substruct-embedding} states that every embedding dimension must represent the count or presence of some ordered subgraph $S_i$ in the embedded graph $G$.
Based on this requirement we now proof the main theorem which shows the relation between \acp{svm} and \ac{lta}.
\begin{thm}\label{thm:ltag:svm-ltag-formulation}
	A binary \ac{svm} classifier that takes substructure component embeddings $\varphi(G)$ as its input has an \ac{lta} formulation with a nontrivial decomposition function $\psi$.
\end{thm}
\begin{proof}
	Let $h: \mathcal{G} \to {\{ -1, +1 \}}$ be a binary graph classifier expressed as $h = h_{\text{\acs*{svm}}} \circ \varphi$ where $\varphi: \mathcal{G} \to \mathbb{N}_0^{d}$ is a substructure component embedding and $h_{\text{\acs*{svm}}}: \mathbb{R}^{d} \to {\{ -1, +1 \}}$ an \ac{svm} classifier.
	By \cref{defn:ltag:substruct-embedding} each $\varphi(G)[i] > 0$ counts or indicates the presence of some ordered subgraph $S_i$ in $G$;
	more precisely:
	\begin{itemize}
		\item If $\varphi(G)[i] = \gamma_i \mathit{count}_{S_i}(G) > 0$, let $C_i \coloneqq {\{ G[s]\, |\, s \in \mathcal{P}(\mathcal{V}_G) \land G[s] \simeq S_i \}}$ be the subgraphs counted by the $i$-th dimension.
		\item If $\varphi(G)[i] = \gamma_i \mathit{contains}_{S_i}(G) > 0$, let $C_i \coloneqq {\left\{ G\left[ \bigcup \{ s \in \mathcal{P}(\mathcal{V}_G)\, |\, G[s] \simeq S_i \} \right] \right\}}$ be the union graph of all subgraphs of $G$ that are isomorphic to $S_i$.
	\end{itemize}
	By interpreting each $C_i$ as a set of isomorphic constituents, we can define the graph decomposition function $\psi_{\varphi}(G) \coloneqq \bigcup_{i=1}^{d} C_i$.

	Based on this decomposition we now bring the \ac{svm} graph classifier $h$ into an \ac{lta} formulation.
	If $h$ is trained on a dataset $\mathcal{D}_{\text{train}} = {\{ (G_1, y_1), \dots, (G_N, y_N) \}}$, via the kernel trick it can be expressed as
	{\allowdisplaybreaks\begin{align*}
		h(G) &= \sgn\left( \sum_{j = 1}^{N} \alpha_j y_j {\langle \varphi(G), \varphi(G_j) \rangle} + b \right)
		\quad\text{for some $\alpha \in \mathbb{R}^N$ and $b \in \mathbb{R}$} \\
		&= \sgn\left( {\sum_{i=1}^{d} {
				\underbrace{\varphi(G)[i]}_{{|C_i|} \gamma_i^{-1}}
				\underbrace{\left( \sum_{j = 1}^{N} \alpha_j y_j {\varphi(G_j)}[i] \right)}_{{} \eqqcolon \beta_i}}
			} + b \right)
		 = \sgn\left( {\sum_{i=1}^{d} {{|C_i|} \gamma_i^{-1} \beta_i}} + b \right) \\
		&= \sgn\left( {\sum_{c \in \psi_{\varphi}(G)} \underbrace{\sum_{i = 1}^d \mathbbm{1}[c \in C_i] \gamma_i^{-1} \beta_i}_{{} \eqqcolon z_i}} + b \right)
		 = \sgn\left( {\sum_{c \in \psi_{\varphi}(G)}} \underbrace{|z_i|}_{{} \eqqcolon w_i} \underbrace{\sgn{z_i}}_{\eqqcolon y_i} + b \right)
		\text{.}
	\end{align*}}
\end{proof}

### 2020-03-14 thesis/content/chapter-related-work.tex
continued svm lta proof + minor fixes

\subsubsection{The 1-dimensional \acs{wl} algorithm}
In the 1-dimensional \ac{wl} algorithm (1-WL), a color is assigned to each vertex of a graph.
If the vertices $v \in \mathcal{V}_G$ of the input graph $G$ are labeled, those labels $l_G[v] \in L_{\mathcal{V}} \subseteq \mathcal{C}$ can be used as the initial graph coloring $\chi_{G,1}^{(0)}(v) \coloneqq l_G[v]$.
Since \ac{wl} colors are inherently discrete, continuous vertex feature vectors $x_G[v]$ are not considered here.
For unlabeled graphs a constant coloring is used, e.g.\ $\forall v \in \mathcal{V}_G: \chi_{G,1}^{(0)}(v) = \textcolor{t_blue}{\texttt{A}}$ for some initial color $\textcolor{t_blue}{\texttt{A}} \in \mathcal{C}$. % chktex 25
In each iteration of the 1-\acs{wl} color refinement algorithm, the following neighborhood aggregation scheme is used to compute a new color for all vertices:
\begin{defn}\label{defn:related:wl1-refine}
	$\chi_{G,1}^{(i+1)}(v) \coloneqq h\left(\chi_{G,1}^{(i)}(v), \ldblbrace \chi_{G,1}^{(i)}(u)\, |\, u \in \Gamma_{G}(v) \rdblbrace\right)$,
	with $\Gamma_G(v)$ denoting the set of adjacent vertices of $v \in \mathcal{V}_G$ and $h: \mathcal{C}^* \to \mathcal{C}$ denoting an injective hash function that assigns a unique color to each finite combination of colors.
\end{defn}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.72\linewidth]{gfx/related-work/wl1-refine.pdf}
	\caption[Example 1-WL color refinement steps.]{
		Example 1-WL color refinement steps.
		After two iterations the coloring stabilizes.
		Each vertex $v$ is labeled with its current color and has its previous color and the colors of the hashed neighbors $\Gamma_G(v)$ written next to it (see \cref{defn:related:wl1-refine}).
	}\label{fig:related:wl1-refine}
\end{figure}
In practice the hash function $h$ is usually defined lazily by using $\mathcal{C} \subseteq \mathbb{N}$ and enumerating color combinations in whichever order they are hashed s.t.\ a new color is introduced every time a previously unseen color combination appears at runtime\footnote{
	Though $\mathcal{C}$ is formally countably infinite, we assume it only contains the finite no.\ of colors in a dataset.
}.

>>>>

\subsubsection{The 1-dimensional \acs{wl} algorithm}
In the 1-dimensional \ac{wl} algorithm (1-WL), a color is assigned to each vertex of a graph.
If the vertices $v \in \mathcal{V}_G$ of the input graph $G$ are labeled, those labels $l_G[v] \in L_{\mathcal{V}} \subseteq \mathcal{C}$ can be used as the initial graph coloring $\chi_{G,1}^{(0)}(v) \coloneqq l_G[v]$.
Since \ac{wl} colors are inherently discrete, continuous vertex feature vectors $x_G[v]$ are not considered here.
For unlabeled graphs a constant coloring is used, e.g.\ $\forall v \in \mathcal{V}_G: \chi_{G,1}^{(0)}(v) = \textcolor{t_blue}{\texttt{A}}$ for some initial color $\textcolor{t_blue}{\texttt{A}} \in \mathcal{C}$. % chktex 25
In each iteration of the 1-\acs{wl} color refinement algorithm, the following neighborhood aggregation scheme is used to compute a new color for all vertices:
\begin{defn}\label{defn:related:wl1-refine}
	$\chi_{G,1}^{(i+1)}(v) \coloneqq h\left(\chi_{G,1}^{(i)}(v), \ldblbrace \chi_{G,1}^{(i)}(u)\, |\, u \in \Gamma_{G}(v) \rdblbrace\right)$,
	with $\Gamma_G(v)$ denoting the set of adjacent vertices of $v \in \mathcal{V}_G$ and $h: \mathcal{C}^* \to \mathcal{C}$ denoting an injective hash function that assigns a unique color to each finite combination of colors.
\end{defn}
\begin{figure}[ht]
	\centering
	\includegraphics[width=0.72\linewidth]{gfx/related-work/wl1-refine.pdf}
	\caption[Example 1-WL color refinement steps.]{
		Example 1-WL color refinement steps.
		After two iterations the coloring stabilizes.
		Each vertex $v$ is labeled with its current color and has its previous color and the colors of the hashed neighbors $\Gamma_G(v)$ written next to it (see \cref{defn:related:wl1-refine}).
	}\label{fig:related:wl1-refine}
\end{figure}
In practice the hash function $h$ is usually defined lazily by using $\mathcal{C} \subseteq \mathbb{N}$ and enumerating color combinations in whichever order they are hashed s.t.\ a new color is introduced every time a previously unseen color combination appears at runtime\footnote{
	Though $\mathcal{C}$ is countably infinite, we assume it only contains the finite no.\ of colors in a dataset.
}.
